Dissecting the interaction between skill level and mental templates in Late Acheulean handaxe morphology: Archaeological and experimental insights

### 2022-11-06 manuscript/manuscript.Rmd
minor revisions.

---
title: "Detecting skill level and mental templates in Late Acheulean handaxe morphology: Archaeological and experimental insights"
author: 
- Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; raylc1996@outlook.com]
- Nada Khreisheh^[The Ancient Technology Centre, Cranborne, Dorset, UK; nada.khreisheh@dorsetcouncil.gov.uk]
- Dietrich Stout^[Department of Anthropology, Emory University, Atlanta, GA, USA; dwstout@emory.edu]
- Justin Pargeter^[Department of Anthropology, New York University, New York, NY, USA; Palaeo-Research Institute, University of Johannesburg, Auckland Park, South Africa; justin.pargeter@nyu.edu]
abstract: "Despite the extensive literature focusing on Acheulean handaxes, especially the sources and meaning of their morphological variability, many aspects of this topic remain elusive. Archaeologists cite many factors that contribute to the considerable variation of handaxe morphology, including knapper skill levels and mental templates. Here we present results from a multidisciplinary study of Late Acheulean handaxe-making skill acquisition involving thirty naïve participants trained for up to 90 hours in Late Acheulean style handaxe production and three expert knappers. We compare their handaxe to the Late Acheulean handaxe assemblage from Boxgrove, UK. Through the principal component analysis of morphometric data derived from images, our study suggested that knapper skill levels and mental templates have a relatively clear manifestation in different aspects of handaxe morphology. The former relates to cross-sectional thinning (PC1), while the latter refers to handaxe elongation and pointedness (PC2). Moreover, we also evaluated the effects of training using the data from a 90-hour-long knapping skill acquisition experiment. We found that reaching the skill level of modern experts requires more training time than was permitted in this extensive and long-running training program. \\par \\textbf{Keywords:} Late Acheulean; Handaxe production; Boxgrove; Experimental archaeology; Skill level; Mental template"
fontsize: 11pt
output: bookdown::pdf_document2
header-includes:
   - \usepackage{textcomp}
   - \usepackage{utopia}
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{float}
   - \floatplacement{figure}{H}
   - \usepackage{caption}
   - \captionsetup[figure]{font=scriptsize}
   - \captionsetup[table]{font=scriptsize}
linestretch: 1.5
link-citations: yes
linkcolor: blue
bibliography: bibliography.bib
csl: apa.csl
---

@@

The morphological variability of Acheulean handaxes has been one of the most well-studied and well-published topics in paleolithic archaeology [@white1998; @earlyhu1998; @key2019]. Despite the recurrent narrative emphasizing the homogeneity and longevity of handaxe assemblages on a global scale and the conservatism behind this phenomenon that evokes genetic explanations [@corbey2016; @corbey2020; @richerson2005; @sterelny2004], many researchers have recognized the diversity within what has been deemed as a unified Acheulean "tradition" and tried to dissect the sources and meaning of this variation [@lycett2008; @nowell2010; @nowell2002; @sharon2011]. More specifically, a complex suite of interconnecting factors have been identified to contribute to handaxe morphological variation, including but not limited to raw material variability [@sharon2008; @eren2014; @lycett2016], percussor properties [@shipton2009], functional differences [@machin2007; @key2016; @kohn1999; @white2018; @key2017], reduction method/intensity [@shipton2009; @shipton2015], learning processes [@lycett2016; @kempe2012], knapper skill levels [@stout2014; @herzlinger2017; @caruana2021], and mental templates [@garcía-medrano2019; @hutchence2021]. From this extensive list, knapper skill levels and mental templates have been repeatedly mentioned and discussed in the now extensive corpus of handaxe studies, and Boxgrove handaxes have been one of the most studied assemblages from these two angles. Of particular attention here are the experimental works conducted by Stout et al. [-@stout2014] focusing on inferring high knapping skill level and Garcia-Medrano et al. [-@garcía-medrano2019] identifying the mental template of the Boxgrove assemblage. Our paper combines these two perspectives and provides novel insights to the same archaeological assemblage by comparing it with experimentally made handaxes.

@@

This work was supported by funding from the National Science Foundation of the USA (grants SMA-1328567 & DRL-1631563), the John Templeton Foundation (grant 47994), and the Emory University Research Council. The handaxe knapping skill acquisition experiment involved in this study was approved by Emory University's Internal Review Board (IRB study no: 00067237). We would also like to thank Radu Iovita for providing us access to the digital photographs of the Boxgrove handaxe assemblage.

>>>>

---
title: "Dissecting the interaction between skill level and mental templates in Late Acheulean handaxe morphology: Archaeological and experimental insights"
author: 
- Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; raylc1996@outlook.com]
- Nada Khreisheh^[The Ancient Technology Centre, Cranborne, Dorset, UK; nada.khreisheh@dorsetcouncil.gov.uk]
- Dietrich Stout^[Department of Anthropology, Emory University, Atlanta, GA, USA; dwstout@emory.edu]
- Justin Pargeter^[Department of Anthropology, New York University, New York, NY, USA; Palaeo-Research Institute, University of Johannesburg, Auckland Park, South Africa; justin.pargeter@nyu.edu]
abstract: "Despite the extensive literature focusing on Acheulean handaxes, especially the sources and meaning of their morphological variability, many aspects of this topic remain elusive. Archaeologists cite many factors that contribute to the considerable variation of handaxe morphology, including knapper skill levels and mental templates. Here we present results from a multidisciplinary study of Late Acheulean handaxe-making skill acquisition involving thirty naïve participants trained for up to 90 hours in Late Acheulean style handaxe production and three expert knappers. We compare their handaxe to the Late Acheulean handaxe assemblage from Boxgrove, UK. Through the principal component analysis of morphometric data derived from images, our study suggested that knapper skill levels and mental templates have a relatively clear manifestation in different aspects of handaxe morphology. The former relates to cross-sectional thinning (PC1), while the latter refers to handaxe elongation and pointedness (PC2). Moreover, we also evaluated the effects of training using the data from a 90-hour-long knapping skill acquisition experiment. We found that reaching the skill level of modern experts requires more training time than was permitted in this extensive and long-running training program. \\par \\textbf{Keywords:} Late Acheulean; Handaxe production; Boxgrove; Experimental archaeology; Skill level; Mental template"
fontsize: 11pt
output: bookdown::pdf_document2
header-includes:
   - \usepackage{textcomp}
   - \usepackage{utopia}
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{float}
   - \floatplacement{figure}{H}
   - \usepackage{caption}
   - \captionsetup[figure]{font=scriptsize}
   - \captionsetup[table]{font=scriptsize}
linestretch: 1.5
link-citations: yes
linkcolor: blue
bibliography: bibliography.bib
csl: apa.csl
---

@@

The morphological variability of Acheulean handaxes has been one of the most well-studied and well-published topics in paleolithic archaeology [@white1998; @earlyhu1998; @key2019]. Despite the recurrent narrative emphasizing the homogeneity and longevity of handaxe assemblages on a global scale and the conservatism behind this phenomenon that evokes genetic explanations [@corbey2016; @corbey2020; @richerson2005; @sterelny2004], many researchers have recognized the diversity within what has been deemed as a unified Acheulean "tradition" and tried to dissect the sources and meaning of this variation [@lycett2008; @nowell2010; @nowell2002; @sharon2011]. More specifically, a complex suite of interconnecting factors have been identified to contribute to handaxe morphological variation, including but not limited to raw material variability [@sharon2008; @eren2014; @lycett2016], percussor properties [@shipton2009], functional differences [@machin2007; @key2016; @kohn1999; @white2018; @key2017], reduction method/intensity [@shipton2009; @shipton2015], learning processes [@lycett2016; @kempe2012], time budgets [@schillinger2014], knapper skill levels [@stout2014; @herzlinger2017; @caruana2021], and mental templates [@garcía-medrano2019; @hutchence2021]. From this extensive list, knapper skill levels and mental templates have been repeatedly mentioned and discussed in the now extensive corpus of handaxe studies, and Boxgrove handaxes have been one of the most studied assemblages from these two angles. Of particular attention here are the experimental works conducted by Stout et al. [-@stout2014] focusing on inferring high knapping skill level and Garcia-Medrano et al. [-@garcía-medrano2019] identifying the mental template of the Boxgrove assemblage. Our paper combines these two perspectives and provides novel insights to the same archaeological assemblage by comparing it with experimentally made handaxes.

@@

We would like to thank Thomas Jennings and three other anonymous reviewers for their insightful feedback on an earlier draft of this manuscript. This work was supported by funding from the National Science Foundation of the USA (grants SMA-1328567 & DRL-1631563), the John Templeton Foundation (grant 47994), and the Emory University Research Council. The handaxe knapping skill acquisition experiment involved in this study was approved by Emory University's Internal Review Board (IRB study no: 00067237). We would also like to thank Radu Iovita for providing us access to the digital photographs of the Boxgrove handaxe assemblage.
